
```{r setoptions, echo=TRUE}
require(knitr)
opts_chunk$set(echo = TRUE, cache = TRUE, cache.path = "cache/", fig.path = "figure/")

```

## Summary


This document investigates daily activity in terms of steps taken within five minute intervals.  The average and median are considered to determine how the data is distributed.  In addition, daily patterns are addressed by average steps per interval frequencies for all days and days broken up by weekend and weekday.



## Loading and preprocessing the data

```{r echo=TRUE}
#required libraries
library(dplyr)
library(lattice)

#load data
activity<-read.table("activity.csv", header=TRUE, stringsAsFactor=FALSE, sep=",")

#omit NA values
activity_wo_na<-activity[!is.na(activity$steps),]

```







## What is mean total number of steps taken per day?

The following shows the frequency of daily activity in average number of steps.   The data was grouped by five minute intervals for each day, and then the average was found for each of these intervals.

```{r echo=TRUE}
#group by day and sum steps taken
grouped_by_day<- group_by(activity_wo_na, date)
summary_steps <- summarise(grouped_by_day,total = sum(steps))

#find the mean and median
avgsteps<-mean(summary_steps$total)
mediansteps<-median(summary_steps$total)

#Construct a histogram with mean line
hist(summary_steps$total, xlab="Total Number of Steps", main="Frequency of Total Number of Steps per Day", col=8, breaks=12)
abline(v = avgsteps, col = 2, lwd = 2)
  
```

The average daily activity is `r avgsteps` steps with median `r mediansteps`.  The median and mean are very similar with the NA values have been dropped.  Therefore, the data is evenly distributed around the mean.



## What is the average daily activity pattern?
```{r echo=TRUE}
#Group by intervals and find average
grouped_by_int<- group_by(activity_wo_na, interval)
summary_stepsint <- summarise(grouped_by_int,  avg = mean(steps))

#plot
plot(strptime(sprintf("%04d", summary_stepsint$interval), format="%H%M"), summary_stepsint$avg, type="l", xlab="Time of Day")

#Information on max steps
summary_stepsint[summary_stepsint$avg==max(summary_stepsint$avg),]
```

There is a significant spike in average acitivity in early morning and fairly large average steps through the late evening.   During the hours of 22:00 - 5:00 there is relatively no activity.  The highest average number of steps for all days occurs at 8:35AM.




## Imputing missing values

The previously omitted missing values will be replaced with the average for the 5 minute interval in which the NA occurred.  This replacement was chosen in order to replace missing data with the most accurate apprximately for the relevant interval.

The previous calculations to produce the histogram and average/median are then repeated.

```{r echo=TRUE}
#copy original data
activity_na_replace<-activity

#copy the intervals to an individual vector to loop over
M<-summary_stepsint$interval
for(i in M) 
  {
#The average for this interval
  int_avg<-summary_stepsint[summary_stepsint$interval==i,]$avg
  
#if there are entries which are NA for this interval, replace them with the interval average  
	if(dim(activity_na_replace[activity_na_replace$interval==i & is.na(activity_na_replace$steps),])[1]>0)
		{
activity_na_replace[activity_na_replace$interval==i & is.na(activity_na_replace$steps),]$steps = int_avg
		}		 			 
	}


#group by day the new data set, find mean/median, and plot 
grouped_by_day<- group_by(activity_na_replace, date)
summary_steps <- summarise(grouped_by_day,
  total = sum(steps))
  
avgsteps<-mean(summary_steps$total)
mediansteps<-median(summary_steps$total)

hist(summary_steps$total, xlab="Total Number of Steps", main="Frequency of Total Number of Steps per Day", col=8, breaks=12)
abline(v = avgsteps, col = 2, lwd = 2)

```

The average daily activity is `r avgsteps` steps with median `r mediansteps`.  The result of using the interval average is the mean and median for the average steps per day be the same.


## Are there differences in activity patterns between weekdays and weekends?

The following data is presented twice.  Once as panels per the assignment description, and once as a superimposed graph.  I included both, because I find the superimposed version easier for visual comparisons.



```{r echo=TRUE}
#create a factor that distinguishes weekday and weekend
activity_na_replace$wday<-as.character(wday(activity_na_replace$date))
activity_na_replace[activity_na_replace$wday==1 | activity_na_replace$wday==6,]$wday = "Weekend"
activity_na_replace[activity_na_replace$wday!="Weekend",]$wday = "Weekday"
activity_na_replace$wday<-as.factor(activity_na_replace$wday)

#group data by interval and weekend/weekday
grouped_by_int<- group_by(activity_na_replace, interval, wday)
 summary_stepsint <- summarise(grouped_by_int,  avg = mean(steps))

#Plot the data
int<- as.POSIXct(strptime(sprintf("%04d", summary_stepsint$interval), format="%H%M"))

xyplot(summary_stepsint$avg~int|summary_stepsint$wday,type="l", auto.key=TRUE, scales=list(x=list(at=NULL)), xlab="Time (Between 00:00 and 23:55)", lwd=2)

xyplot(summary_stepsint$avg~int, groups =summary_stepsint$wday,type="l", auto.key=TRUE, scales=list(x=list(at=NULL)), xlab="Time (Between 00:00 and 23:55)", lwd=2)

#split data for weekend and weekdays
wdays<-summary_stepsint[summary_stepsint$wday=="Weekday",]
wend<-summary_stepsint[summary_stepsint$wday=="Weekend",]

#summary for weekend/weekday
summary(wdays)
summary(wend)

#Find the interval where the max occurs
wdays[wdays$avg==max(wdays$avg),]
wend[wend$avg==max(wend$avg),]
```

From the graph, one can see that the both weekend and weekdays have spikes in early morning;  however general activity starts earlier on weekday mornings.  Weekend afternoons also have more activity on average than weekday afternoons.  There is also some extra late activity on weekends.  More specifically, once see that the highest average of steps for the weekdays remains at 8:35AM.  However, this occurs slightly later - 8:50AM - on weekends.  However, the average number of steps for these intervals are approximately 211 in both cases.

